A Plasma Torus Around a Young Low-Mass Star

### 2025-03-13 paper/ms.tex
update

Age                                     & Adopted stellar age (Myr)\dotfill                              & $16^{+19}_{-6}$  & 2 \\

@@

2: Gaia DR3 empirical isochrone fit (see~\ref{subsec:stparams}).

@@

The other potential analog for the CPVs are the $\sigma$~Ori~E
variables, a rare subset of B stars with radiatively-driven winds
that accumulate into
warped plasma tori \cite{Townsend2005,Townsend2008}.  These tori tend
to have dense antipodal accumulations of plasma sculpted by
tilted-dipole magnetic fields, and the transits of
these clumps are thought to produce the observed broadband photometric variability
through bound-free scattering \cite{Townsend2005} and
Thomson scattering \cite{Berry2022}.  For $\sigma$~Ori~E and almost
all of its analogs, the result is light curves that appear ``simple'',
resembling those of eclipsing binaries \cite{Townsend2008}.  The two known
exceptions, HD~37776 and HD~64740, show complex light curves
resembling CPVs \cite{Mikulasek2020,Bouma2024} and have
spectropolarimetric magnetic field maps indicating strong
contributions from higher-order magnetic moments
\cite{Kochukhov2011,Shultz2018}.  There are two implications: first,
the complexity of CPVs may be a direct consequence of magnetic fields
with highly multipolar contributions.  Second, CPVs could be a
source of astrophysical false positives in photometric searches for
eclipsing binaries and transiting exoplanets around young
pre-main-sequence M dwarfs \cite{Johns-Krull2016,Bouma2020}.

@@

{\it Age: Isochrones}---The color and absolute magnitude of
TIC~141146667 suggest that it is a pre-main-sequence M dwarf, similar
to all other known CPVs \cite{Stauffer2017,Stauffer2021,Bouma2024}.
The star's proximity ($d$=58\,pc) and its high galactic latitude
($b$=$+$53$^\circ$) yield negligible interstellar reddening along the
line of sight \cite{Green2019}.
{\bf Figure~X} shows the location of the star in the color--absolute
magnitude diagram (CAMD)
relative to young stellar populations including Upper Scorpius (USco),
Upper-Centaurus-Lupus (UCL), IC~2602, and the Pleiades.
To make this diagram, {\bf TODO}we adopted the USco and UCL members
from X, the IC~2602 members from Y, and the Pleiades members from Z,
and applied the same reddening corrections as in \cite{Bouma2022}.
To measure the isochrone age of TIC~141146667, we then followed the 
empirical approach proposed by \cite{Gagne2020} and
implemented in \cite{Bouma2022}.
The idea is to estimate the star (or star cluster's) age by
interpolating between the observed isochrones of star clusters with
ages that have been otherwise calibrated, for instance using
the lithium depletion-boundary method.
While this circumvents known modelling issues in molecular opacities
and starspot coverage for young M dwarfs,
the ages returned by this procedure depend on the ages
assumed for each reference cluster.
In this case, the most important reference is UCL, since TIC~141146667
overlaps directly on its isochrone on the CAMD.
We adopted the 16~Myr age for UCL from \cite{PecautMamajek2016}, which
is consistent with ages determined by more recent sub-clustering
analyses \cite{Ratzenbock2023} for the largest sub-populations.
This in turns yields an isochronal age for TIC~141146667 of
$t_{\rm iso}$=$16^{+19}_{-6}$\,Myr.
Qualitatively, this range of uncertainty is consistent with no stars
in the $\approx$40\,Myr IC~2602 cluster with similar redness as
TIC~141146667 being equally bright.
%FIXME: update discussion following consideration of anomalous CPV
% brightnesses and colors...

@@

@@

{\it Age: Summary}---The main indicators for the youth of
TIC~141146667 are {\it i)} that it is a complex periodic variable, and
{\it ii)} that it is 1.5 magnitudes brighter than main sequence stars
of the same color, while showing no indicators for binarity.  The
first item constitutes evidence for youth since all known CPVs are
$\lesssim$200\,Myr \cite{Bouma2024}.  Reference sequences for
$\approx$40-50\,Myr clusters show no stars of equal luminosity and the
same temperature, suggesting a photometric isochronal age upper limit
of 50\,Myr.  The weak lithium absorption suggests an age of at least
20\,Myr based on an empirical comparison using Gaia-ESO spectra, or at
least 35\,Myr based on the \cite{Feiden2016} isochrones.  These
considerations yield our adopted age range of 20-50\,Myr.

>>>>

Age                                     & Stellar age range (Myr)\dotfill                                & 35-150  & 2 \\

@@

2: Gaia DR3 photometry and lack of lithium (see~\ref{subsec:stparams}).

@@

The other potential analog for the CPVs are the $\sigma$~Ori~E
variables, a rare subset of B stars with radiatively-driven winds
that accumulate into
warped plasma tori \cite{Townsend2005,Townsend2008}.  These tori tend
to have dense antipodal accumulations of plasma sculpted by
tilted-dipole magnetic fields, and the transits of
these clumps are thought to produce the observed broadband photometric variability
through bound-free scattering \cite{Townsend2005} and
Thomson scattering \cite{Berry2022}.  For $\sigma$~Ori~E and almost
all of its analogs, the result is light curves that appear ``simple'',
resembling those of eclipsing binaries \cite{Townsend2008}.  The two known
exceptions, HD~37776 and HD~64740, show complex light curves
resembling CPVs \cite{Mikulasek2020,Bouma2024} and have
spectropolarimetric magnetic field maps indicating strong
contributions from higher-order magnetic moments
\cite{Kochukhov2011,Shultz2018}.  There are two implications: first,
the complexity of CPVs may be a direct consequence of magnetic fields
with highly multipolar contributions.  Second, CPVs could be a
source of astrophysical false positives in photometric searches for
eclipsing binaries and transiting exoplanets around young
pre-main sequence M dwarfs \cite{Johns-Krull2016,Bouma2020}.

@@

{\it Age: Isochrones}---The color and absolute magnitude of
TIC~141146667 suggest that it is a pre-main sequence M dwarf, similar
to all other known CPVs \cite{Stauffer2017,Stauffer2021,Bouma2024}.
The star's proximity ($d$=58\,pc) and its high galactic latitude
($b$=$+$53$^\circ$) yield negligible interstellar reddening along the
line of sight \cite{Green2019}.  Figure~\ref{fig:camd} shows the
location of the star in the color--absolute magnitude diagram (CAMD)
relative to young stellar populations including Upper Scorpius (USco),
IC~2602, and the Pleiades.  To make this diagram, we adopted the USco
members in the $\delta$~Sco and $\sigma$~Sco sub-associations from
\cite{Ratzenbock2023}, and the IC~2602 and Pleiades members from
\cite{Hunt2024}.  We assumed an average $V$-band extinction $A_{\rm
V}$=${0.12, 0.11, 0.10}\,{\rm mag}$ for USco \cite{Pecaut2016},
IC~2602 \cite{Hunt2024}, and the Pleiades \cite{Hunt2024}
respectively.  We dereddened the photometry using the extinction
coefficients $k_X\equiv A_X/A_0$ tabulated in
\citet{GaiaCollaboration2018}, assuming that $A_0 = 3.1 E(B-V)$.

Figure~\ref{fig:camd} suggests that TIC~141146667 is
older than USco, and potentially younger than the Pleiades.  The precision of this
statement is set by the intrinsic scatter of the calibration
sequences; the most luminous stars in the Pleiades of the same color
have a similar absolute magnitude as TIC~141146667.
Previous work \cite{Stauffer2021} has noted that in the Gaia
passbands, CPVs tend to be photometrically redder and more luminous
than single stars in any given cluster.
While this effect complicates any attempt at age inference based on this
photometry, it suggests that the Pleiades may be a better comparison
population than IC~2602.
Broadly, we take the star's isochrone location to suggest age bounds
$t_{\rm CAMD}$$\sim$40-150\,Myr.

%  and applied the same reddening corrections as in \cite{Bouma2022}.
%  To measure the isochrone age of TIC~141146667, we then followed the 
%  empirical approach proposed by \cite{Gagne2020} and
%  implemented in \cite{Bouma2022}.
%  The idea is to estimate the star (or star cluster's) age by
%  interpolating between the observed isochrones of star clusters with
%  ages that have been otherwise calibrated, for instance using
%  the lithium depletion-boundary method.
%  While this circumvents known modelling issues in molecular opacities
%  and starspot coverage for young M dwarfs,
%  the ages returned by this procedure depend on the ages
%  assumed for each reference cluster.
%  In this case, the most important reference is UCL, since TIC~141146667
%  overlaps directly on its isochrone on the CAMD.
%  We adopted the 16~Myr age for UCL from \cite{PecautMamajek2016}, which
%  is consistent with ages determined by more recent sub-clustering
%  analyses \cite{Ratzenbock2023} for the largest sub-populations.
%  This in turns yields an isochronal age for TIC~141146667 of
%  $t_{\rm iso}$=$16^{+19}_{-6}$\,Myr.
%  Qualitatively, this range of uncertainty is consistent with no stars
%  in the $\approx$40\,Myr IC~2602 cluster with similar redness as
%  TIC~141146667 being equally bright.
%FIXME: update discussion following consideration of anomalous CPV
% brightnesses and colors...

@@

\begin{figure}[!t]
  \centering
  \includegraphics[width=0.7\textwidth]{figures/sf3.pdf}
  \caption{Dereddened Gaia DR3 color vs.~absolute magnitude diagram of
  TIC~141146667 and comparison samples. 
  TIC~141146667 is on the pre-main sequence; stars with the same color
  that have arrived at the main sequence are typically 1.5\,magnitudes
  fainter.  The star's location in this diagram broadly suggests an
  age of 40-110\,Myr.  }
  \label{fig:camd}
\end{figure}

@@

{\it Age: Summary}---The main indicators for the youth of
TIC~141146667 are {\it i)} that it is a complex periodic variable, and
{\it ii)} that it is 1.5 magnitudes brighter than main sequence stars
of the same color, while showing no indicators for binarity.  The
first item constitutes evidence for youth since all known CPVs are
$\lesssim$200\,Myr \cite{Bouma2024}.  The reference sequences for the
Pleiades (112\,Myr, \cite{Dahm2015}) show a few stars of equal
luminosity and the same temperature, suggesting a photometric
isochronal age upper limit $\lesssim$150\,Myr.  The weak lithium
absorption suggests an age of at least 20\,Myr based on an empirical
comparison using Gaia-ESO spectra, or at least 35\,Myr based on the
\cite{Feiden2016} isochrones.  These considerations yield our adopted
age range of 35-150\,Myr.
